---
title: "Basic survival analysis in R"
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

---

### Getting started

The [survival package](https://cran.r-project.org/web/packages/survival/index.html) is installed along with the `base` functions in R, so all you need to do after opening R is to load the package using the `library()` function.

The package comes with various in built datasets. For a full list, see the start of the index of [the survival reference manual](https://cran.r-project.org/web/packages/survival/survival.pdf). Each dataset is described in detailed in the manual. 

Let us load the survival package and take a look at the dataset `myeloid`. 

This is a simulated dataset based on a trial in acute myeloid leukemia, containing 646 observations of 9 variables. Among these variables, `futime` is the time to death or last time of follow-up, `death` is an event indicator (equal to 1 if the time listed in `futime` is a death and 0 if it is a censoring event), `trt` is the treatment arm (either arm A or arm B) and `sex` is the sex (where "f" is female and  "m" is male). 

The function `head()` show the first 6 rows of the dataset.

```{r starting}
library(survival)
head(myeloid)
```

---

### Kaplan-Meier survival curves

Survival curves can be plotted using the `survfit()` function, which take a formula as an argument.

In the survival package the specification of time-to-event outcomes on the right side of the equation in the formula is done using the `Surv()` function. This function take the variable name used for the individual's event times and event indicator(s) as arguments.

In the `myeloid` dataset the event times was called `futime` and the event indicators was called `death`, and `death==1` corresponded to an event time.

The below code will produce a plot of the Kaplan-Meier survival function.

```{r kaplan-meier, fig.align="center"}
km <- survfit(Surv(futime, death==1) ~ 1, myeloid)
plot(km, xlab="Time", ylab="Survival probability")
```

---

Pointwise 95\% confidence intervals are added by default when only a single curve is plotted (which is the case when the right side of th formula is `1` as above).

For more options see for example the R help file for `plot.survfit` by typing `?plot.survfit` in R. 

Remember that for all R commands useful information on options, and often runnable examples, are available in the help files that can be accessed by typing `?` and the name of the command.

### Nelson-Aalen cumulative hazard curves

Similarly, a plot of the corresponding Nelson-Aalen cumulative hazard function can be produced by using the option `cumhaz=TRUE` when plotting the `survfit` object.

```{r nelson-aalen, fig.align="center"}
plot(km, cumhaz=T, xlab="Time", ylab="Cumulative hazard")
```

---

### The log-rank test

When interested in comparing survival curves between groups, these can be plotted by adding a categorical grouping variable on the right side of the equation in `survfit`. One such variable is the `trt` variable in the `myeloid` dataset.

Se the below code, which produce separate curves for treatment group A and B, with an added legend using the `legend()` function.

```{r kaplan-meier2, fig.align="center"}
km2 <- survfit(Surv(futime, death==1) ~ trt, myeloid)
plot(km2, col=c(1, 2), xlab="Time", ylab="Survival probability")
legend("topright", c("trt = A", "trt = B"), col=1:2, lwd=2)
```

The most common test for comparing two survival curves is the log-rank test. This is here implemented with the `survdiff` function, and can be used as shown in the below code.

```{r log-rank, fig.align="center"}
survdiff(Surv(futime, death==1) ~ trt, myeloid)
```

We see that the p-value for the test is 0.002, which indicates a significant difference between groups at a 5\% significance level.

---

### The Cox proportional hazard model

A similar test between groups can also be performed using a Cox proportional hazard model, with the function `coxph` as shown below.

```{r cox, fig.align="center"}
cfit <- coxph(Surv(futime, death==1) ~ trt, myeloid)
summary(cfit)
```

We here see a p-value of 0.00206 for difference between groups, and an estimated hazard ratio of 0.7077 in favour of treatment B.

More independent variables can now be included by adding variables to the right side of equation. The below code show a Cox model with `trt` and `sex` as independent variables.

```{r cox2, fig.align="center"}
cfit <- coxph(Surv(futime, death==1) ~ trt + sex, myeloid)
summary(cfit)
```

---

### Further references

For a more thorough introduction, see for example [the survival package vignette](https://cran.r-project.org/web/packages/survival/vignettes/survival.pdf) and the other vignettes of the [survival package](https://cran.r-project.org/web/packages/survival/index.html).

</br>
